---
title: "Statistical Inference Analysis on ToothGrowth Dataset"
author: "Dhruv Mishra"
output: word_document
---
The `ToothGrowth` data set contains "*The response is the length of odontoblasts (teeth) in each of 10 guinea pigs at each of three dose levels of Vitamin C (0.5, 1, and 2 mg) with each of two delivery methods (orange juice or ascorbic acid).*" [Ref.; C. I. Bliss (1952) The Statistics of Bioassay. Academic Press]

### Exploratory data analyses
The values which the dose variable can take are 0.5, 1.0 and 2.0 and thus we will convert it into a factor. The relevant plot is in the appendix section.
```{r, warning=FALSE}
library(ggplot2)
data(ToothGrowth)
ToothGrowth$dose <- as.factor(ToothGrowth$dose)
```

The distributions of tooth length by delivery method and dose levels seem to indicate that for 0.5 and 1 mg, the mean tooth length is greater for those where orange juice (OJ) was used. For the 2 mg level, no noticeable difference in the mean is observed, but the interquartile range  for the OJ group seems to be smaller. 

### Basic Summary of data
```{r}
summary(ToothGrowth)
```

### Confidence intervals and hypothesis tests
See if the difference if mean equal to zero is present in the 95 percent confidence interval or t statistic is significantly larger or p value is very large than 0.05. The relevant codes are in the appendix section.

##### Supp OJ vs VC
The confidence interval contains zero so we can't reject the null hypothesis which means that we can't say that the delivery method has an effect on the tooth length.

##### Dose 0.5 vs 1
The confidence interval of -11.98 to -6.27 means that the null hypothesis is rejected and the growth caused by 0.5 is less that 1. The p value is also very small implying the rejection of null hypothesis.

##### Dose 1 vs 2
The confidence interval of -8.99 to -3.73 means that the null hypothesis is rejected and the growth caused by 1 is less that 2. The p value is also very small implying the rejection of null hypothesis.

The above test implies the order 2>1>0.5 for dose variable.

### Assumptions
For this experiment, we've assumed there's a common variance in the guinea pigs population (var.equal=TRUE) while performing tests on the dose variable. The experiment design suggest that each Guinea pig was randomly assigned to a combination of dosage and supplement type, so the test that we performed used the independent samples methodology.

### Conclusion
**The supplement type has no effect on tooth growth. However increasing the dose level leads to increased tooth growth.**

### Appendix
#### Exploratory data analysis
```{r}
ggplot(data=ToothGrowth, aes(y=len, x=supp, fill=supp)) + geom_boxplot() + facet_wrap(~ dose, ncol=3) + ylab("Tooth length") + xlab("Delivery method") + ggtitle("Tooth growth by delivery and dose") + stat_summary(fun.y=mean, geom="point", shape=7, size=2)
```


The square dot represents the respective mean.

#### Hypothesis Tests
##### Supp OJ vs VC
```{r}
t.test(ToothGrowth$len[ToothGrowth$supp=="OJ"], ToothGrowth$len[ToothGrowth$supp=="VC"], paired = FALSE, var.equal = FALSE)
```

##### Dose 0.5 vs 1
```{r}
t.test(ToothGrowth$len[ToothGrowth$dose==0.5], ToothGrowth$len[ToothGrowth$dose==1], paired = FALSE, var.equal = TRUE)
```

##### Dose 1 vs 2
```{r}
t.test(ToothGrowth$len[ToothGrowth$dose==1], ToothGrowth$len[ToothGrowth$dose==2], paired = FALSE, var.equal = TRUE)
```